---
title: "Instructions and Background Information"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
always_allow_html: true
output:
  html_fragment:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results='asis', echo=FALSE, warning=FALSE, message = FALSE)
# needed for trouble shooting
boo_DEBUG <- FALSE
if(boo_DEBUG==TRUE){
  # myConfig <- file.path(system.file(package="ContDataQC"), "extdata", "config.ORIG.R")
  # source(myConfig)
}## IF ~ boo_DEBUG ~ END
```

The following instructions are separted into two major sections. The first section describes how to operate the "Calculator" tab and the second section the "Site Class Identifier" tab.

# IBI Calculator Instructions

IDEMtools was developed to calculate diatom metrics and Indices of Biotic Integrity (IBI) for perrenial, wadeable streams in Indiana. Detailed documentation of IBI development can be downloaded [here](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IDEMtools/Reports). Users can run any data through the IBI calculator and get a result. However, if samples do not meet the criteria listed on the Background page, results should be interpreted with caution. 

The Instructions are divided into three sections: 1) preparing the input file; 2) app operation; and 3) frequently asked questions (FAQ). Links to an example input file and a document containing more detailed information on preparing the input file are also provided.

This app was developed by Ben Block from Tetra Tech, Inc. (Ben.Block@tetratech.com), with underlying R code written by Ben Block and Erik Leppo (Erik.Leppo@tetratech.com). Please contact Ben Block should any issues or questions arise.

### Foreword
Tetra Tech, Inc. developed the IDEMtools shiny app for the Indiana Department of Environmental Management (IDEM). It allows IDEM and their partners to calculate IBI scores for diatom samples. The diatom IBIs improve IDEM’s diagnostic ability to identify degradation in biological integrity and water quality. The app is streamlined and easy to operate, and only requires an input dataset to function. Shiny apps are interactive web applications that serve as graphical user interfaces. They are linked to R, which is an open source programming language and software environment for statistical computing.

### Preparing the input file
IDEMtools requires input files to be in a tabular format, saved as either comma-separated values (CSV) (most commonly used) or tab-separated (.txt or .tsv) files. Table 1 contains a list of required fields. Column names must be capitalized and match the spelling and symbology exactly as shown (some column headings have underscores).

Click here to view an [EXAMPLE INPUT FILE](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IDEMtools/Example_Files).

If the user is unsure how to download a data file from GitHub, [watch this short video](https://www.youtube.com/watch?v=GIJdfuAoqFI) or contact Ben Block (ben.block@tetratech.com).

The input file is now ready to be run through the IBI calculator!

#### Table 1. List of required fields. Column names must be capitalized and match the spelling and symbology exactly as shown.
```{r table_1}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab1.dir <- "Instr_Table_Diatoms"

table1 <- read_excel(file.path(table.dir, table.file), sheet = tab1.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  

# kable(table1)
table1 %>% 
  kbl() %>% 
  kable_styling(full_width = F, position = "left")
```


### Instructions

Once the 'Calculator' tab is opened, the user will see the IBI calculator interface. The user should follow the onscreen instructions as follows:

1. Load file
    + **Default settings work well**. If the user is unsure of how to format the data, simply store as a .csv and use the default load settings.
    + An example input file can be downloaded from Github [EXAMPLE LINK](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IDEMtools/Example_Files)
    + Choose Separator. The separator indicates how the data is stored. Comma indicates the input file is a comma-separated file (.csv). Tab indicates the input file is a tab-separated file (.txt or .tsv). Finally, Semicolon indicates the input file is a semicolon-separated file which is uncommon in the United States but common in western European countries. Be certain that the designated separator is not used in any of the text fields (e.g., for multiple BC_USGS values, separate with a semicolon).
    + Choose Quote. The quote indicates how the data is stored. Double quote is the most common.
    + Choose file to upload. Hit the browse button and search for the input file. 
    + Once uploaded, make sure data look correct in the viewer. If the incorrect Separator or Quote is chosen, the user may receive an error or the data may look incorrect in the viewer. 

2. Calculate IBI
    + Hit the 'Calculate Metric Values and Scores' button.
    + A progress bar will appear in the bottom, right corner of the screen. Results cannot be downloaded (step 3) until the progress bar is completed and disappears.

3. Download results
    + Select the button to download a zip file with inputs and results.
    + Check ‘results_log.txt’ for any warnings or messages. Note, some warnings are automatically generated by R. Feel free to reach out to Ben Block (Ben.Block@tetratech.com) for any questions related to warnings.

4. Use interactive map and plots for data exploration
    + Navigate to the top of the page and click on the “Data Explorer” tab.
    + Prior to metric scores calculation, a map will not be visible.
    + Once metric scores are calculated, a map will become visible.
    + Sites are clustered when zoomed out for increased visibility - zoom in for detail.
    + Sites are color coded by their Index Narrative value - click on a site for more info.
    + The map can be filtered based on INDEX_REGION using the checkbox group in the upper right. 
    + The map extent can be saved using the print function from the drop-down menu in the upper left.
    
### Frequently asked questions and troubleshooting

1. Why am I getting an error saying that I am missing columns even when I am not?
    + You may have incorrectly spelled a given column. Try writing the column in all capital letters. Also, some columns (e.g., INDEX_REGION) require an underscore to separate the two words.

2. Why does my data look strange in the data viewer?
    + You likely have the incorrect Separator or Quote selected. Otherwise, there may be commas in text fields when the comma separator is selected.

3. The IBI calculation is taking forever to calculate, has the app frozen?
    + Even though R works remarkably fast, a large dataset will cause the app to slow down. The estimate is approximately 30 seconds for small datasets; however, this could extend to a few minutes for large datasets.

4. Why is there no map shown in the “Sites and Scores Map” tab?
    + The map requires the results of the IBI calculator to render.
    
# Site Identifier Instructions

### Foreword
This tab allows the user to upload station information and determine which site class (HiN or LoN) a site belongs to. Sites are assigned to site classes using the watershed-scale mean lithological nitrogen content (NWs) from [StreamCat](https://www.epa.gov/national-aquatic-resource-surveys/streamcat-dataset-0). The function also provides Base Flow Index (BFIcat) values which are necessary for metric calculation. See Table 2 for additional information on the NWs and BFIcat StreamCat variables. See Table 3 for necessary input fields.

* Sites with NWs >= 0.089 % are classified as **HiN**
* Sites with NWs < 0.089 % are classified as **LoN**

For more information on the derivation of these site classes, please read the [IDEM Diatom IBI report](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IDEMtools/Reports).

Note, only COMIDs from the state of Indiana can be used. If the user would like COMIDs from other, nearby states, please contact Ben Block (Ben.Block@tetratech.com) and he can update the available list.


#### Table 2. Background information on StreamCat variables
```{r table_2}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab2.dir <- "StreamCatVars"

table2 <- read_excel(file.path(table.dir, table.file), sheet = tab2.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  

# kable(table2)
table2 %>% 
  kbl() %>% 
  kable_styling(full_width = F, position = "left")
```


#### Table 3. List of required fields. Column names must be capitalized and match the spelling and symbology exactly as shown.
```{r table_3}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab3.dir <- "SiteClassInput"

table3 <- read_excel(file.path(table.dir, table.file), sheet = tab3.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  

# kable(table3)
table3 %>% 
  kbl() %>% 
  kable_styling(full_width = F, position = "left")
```


### Site Class Identifier Instructions

Once opened, the user will see the site class identifier interface. The user should follow the onscreen instructions as follows:

1. Load file
    + **Default settings work well**. If the user is unsure of how to format the data, simply store as a .csv and use the default load settings.
    + An example input file can be downloaded from Github [EXAMPLE LINK](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IDEMtools/Example_Files)
    + Choose Separator. The separator indicates how the data is stored. Comma indicates the input file is a comma-separated file (.csv). Tab indicates the input file is a tab-separated file (.txt or .tsv). Finally, Semicolon indicates the input file is a semicolon-separated file which is uncommon in the United States but common in Western-European countries. Be certain that the designated separator is not used in any of the text fields.
    + Choose Quote. The quote indicates how the data is stored. Double quote is the most common.
    + Choose file to upload. Hit the browse button and search for the input file. 
    + Once uploaded, make sure data is correct in the viewer. If the incorrect Separator or Quote is chosen, the user may receive an error or the data may look incorrect in the viewer. 

2. Assign site classes
    + Hit 'Assign sites to class' button.
    + A progress bar will appear in the bottom, right corner of the screen. Results cannot be downloaded (step 3) until the progress bar is completed and disappears.
    
3. Download results
    + Select the button to download a zip file with inputs and results.
    + Check ‘results_log_sci.txt’ for any warnings or messages. Note, some warnings are automatically generated by R. Feel free to reach out to Ben Block (Ben.Block@tetratech.com) for any questions related to warnings.
