---
title: "Background Information"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
always_allow_html: true
output:
  html_fragment:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results='asis', echo=FALSE, warning=FALSE, message = FALSE)
# needed for trouble shooting
boo_DEBUG <- FALSE
if(boo_DEBUG==TRUE){
  # myConfig <- file.path(system.file(package="ContDataQC"), "extdata", "config.ORIG.R")
  # source(myConfig)
}## IF ~ boo_DEBUG ~ END
```

## Background
This R Shiny application calculates numeric Index of Biotic Integrity (IBI) scores based on diatom samples collected by the Indiana Department of Environmental Management (IDEM).

IDEM monitors the conditions of streams and rivers in Indiana to inform management decisions regarding protection, mitigation, and restoration of aquatic life and other designated uses. The Indiana water quality standards (Administrative Code Article 2) describe minimum acceptable conditions in terms of narrative biological criteria. Following recommendations by the U.S. EPA, IDEM developed numeric Indices of Biological Integrity (IBI) for fish and macroinvertebrates. An IBI for diatoms, developed in 2021 and which can be **calculated in this application**, provides a third biological assemblage for assessing attainment the biological criteria. The IBIs consolidate measures of the aquatic community (metrics) that, when combined, indicate the similarity of a biological sample to expected conditions of a well-balanced community. The process by which the metrics are selected and combined in an index follows established and innovative analytical methods of the reference condition approach (Hughes et al. 1986, Bailey et al. 2004). In this approach, biological conditions that are sampled from relatively undisturbed sites and that account for natural variability are set as a standard (or reference) to which other samples are compared (Stoddard et al. 2006). Using metrics to establish the numeric index scale, IBI values that resemble those found in reference sites are determined to meet the expectations for a well-balanced aquatic community. IBI values that are unlike the reference values indicate departure from the acceptable biological conditions and probable impairment of aquatic life uses. 

The diatom IBI calibration project resulted in two indices applicable in two stream types that varied by the concentration of natural lithologic nitrogen (as estimated by USGS StreamCat). Each stream was designated as one of the two site classes, HiN and LoN, and an index was developed for both. Reports detailing the development of the diatom IBIs can be found [here](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IDEMtools/Reports).

## Intended application
The purpose of these indices are to describe the diatom community of Indiana freshwater streams for the purpose of biological condition assessment. A diatom IBI enhances the state of Indiana’s monitoring and assessment strategy by adding a numeric indicator of diatom community structure that could be used to assess aquatic life use in:
* IDEM’s Integrated Report, thus satisfying 305(b) and 303(d) reporting requirements to U.S. EPA.
* Watershed characterization projects which identify critical areas and chemical/physical stressors to the biological communities.
* Identifying improvements in the biological communities following watershed restoration efforts.

In addition, a diatom IBI provides an accurate assessment of ecological effects on a third assemblage in Indiana, which might respond to different stressors than those indicated by the existing fish and macroinvertebrate assemblages, thus improving IDEM’s diagnostic ability to identify causes of degradation in water quality. Diatoms are typically associated with nutrient availability (Charles et al. 2019) and the diatom IBI might be useful to evaluate direct biological responses to excessive nutrients.

Streams sampled by the diatom monitoring program were wadeable, perennial, and have watersheds draining less than 1,000 square miles. All diatom samples used in the analyses were collected by IDEM field staff according to the Technical Standard Operating Procedures for periphyton sampling (IDEM 2018). Periphyton samples are collected during low flow, from late August through October. If the user applies the indices to samples that do not meet the criteria listed above, results should be interpreted with caution. For more information on sampling methodologies and application limits, see the IBI development reports [here](https://github.com/Blocktt/ShinyAppDocuments/tree/main/IDEMtools/Reports).

## Input metrics and scoring formulae
Diatom IBIs were calibrated for two natural stream types - the HiN and LoN site classes. Table 1 shows the input metrics and scoring formulae for the two IBIs. See the **"Site Class Identifier"** tab for more information on site classes and how to assign sites to each site class.

#### Table 1. Metrics and scoring formulae for the IDEM Diatom IBIs. DE = discrimination efficiency, Trend = metric response to stress, 5th = 5th percentile of metric values, and 95th = 95th percentile of metric values. BFI = base flow index as determined by USGS StreamCat (watershed scale).
```{r table_1}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab1.dir <- "DiatomMetrics"

table1 <- read_excel(file.path(table.dir, table.file), sheet = tab1.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  

options(knitr.kable.NA = '')
table1 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```
